Preparing for Potential Closure of European Airspaces due to Re-entering Space Objects

### 2025-05-05 paper.qmd
final submission

format:
  ieee-pdf:
    keep-tex: true
    # fig-pos: 'H'
    journaltype: conference
    papersize: a4
    include-in-header:
      - file: media/preamble.tex
    include-before-body:
      - text: |
          \let\qty\SI
          % NOTE: make siunitx's \qty work without clashing with the one in physics
  ieee-html: 
    html-math-method: mathjax
    include-in-header:
      file: media/siunitx.html
author:
  - id: es
    name: Enrico Spinielli
    affiliations:
      - ref: aiu
    orcid: 0000-0001-8584-9131
    email: enrico.spinielli@eurocontrol.int
    attributes:
      corresponding: true
  - name: Rainer Koelle
    affiliations: 
      - ref: aiu
    orcid: 0000-0003-2810-1404
    email: rainer.koelle@eurocontrol.int
  - name: Quinten Goens
    affiliations: 
      - ref: aiu
    orcid: 0000-0003-2187-1924
    email: quinten.goens@eurocontrol.int
affiliations:
  - id: aiu
    name: EUROCONTROL
    department: Aviation Intelligence Unit
    city: Brussels
    country: Belgium
bibliography: media/references.bib
acronyms:
  insert_loa: false
  fromfile:
    - ./media/acronyms.yml
editor: 
  markdown: 
    wrap: 72
---
```{r}
```

@@

The recent decades observed an increase in the number of space activities.
This resulted in higher numbers of re-entries at the end of the respective mission
lifespan or orbital decay. The risk of collision between uncontrolled
re-entering space debris and aircraft is increasing due to the continual 
growth of flights and of orbiting bodies and their relevant re-entries. 
<!-- why important --> National and international
authorities need to balance safety and economic impact of adopted
measures. This is particularly challenging should a re-entry event be announced on short notice.
Thus, there is a need for effective tools to reduce the uncertainty and increase the safety awareness of uncontrolled re-entries. 
<!-- approach / experiments --> To determine the risk exposure, this paper follows a data-driven approach.
We employ hourly and daily spatial flight densities calculated from trajectories as recorded by
EUROCONTROL's Network Manager and binned with the Uber H3 geospatial
indexing system. Impact risk probabilities are then calculated based on previous
theoretical work. 
<!-- results / conclusions --> This work enhances the
model by using actual flown flight trajectories as a basis for
time-varying flight densities rather than relying on proxies such as the
population density or ADS-B/Mode S message counts. These densities
combined with a buffered area along the eventual path of re-entry, can
provide mid-air impact probabilities to authorities as an aid for
deciding on safety measures such as airspace closures and/or
re-routings.
The findings and the approach are demonstrated on the basis of an expected re-entry from an earlier space mission.\
       IS IT WORTH TO STATE THE DETERMINED RISK?  \
The approach in this paper enables a computing time efficient mechanism to support the re-entry related decision-making.

@@

With forecasts predicting a steadily growing number of flights, see
@fig-forecast, and the modern race to commercialise space, there is a
need to re-evaluate the risks of collision and casualties due to
uncontrolled debris re-entries. The increasing risk landscape poses
challenges in terms of oversight and risk management. The operational and 
economic demand by air transport requires a trade-off with large-scale
airspace and flow management measures to address the inherent safety risk.
There is a growing demand by overseeing authorities for new methodologies and tools to
assess which safety measures are the most adequate to minimise potential
casualties and economic impact.

The increase in space operations does not exist in a vacuum. The
continual growth of air traffic - despite temporal bumps - is evidenced
by the surging number of flights in Europe [@forecastUpdate2025] and
across the world. On-going development/deployment activities of
mega-constellations of satellites in \acr{LEO}, such as the
already active Starlink and OneWeb or planned ones (e.g. Kuiper,
Starshield, Xingwang), will add thousand of orbiting objects with a short
operational life. Some of these will inevitably re-enter the Earth's
atmosphere in an uncontrolled manner, c.f. [@tbl-megaconstellations;
@megaconstellations].
These numbers and occurrence rates will pose challenges to the steadily increasing air traffic.

@@

Aviation safety risk management is a well-established domain in air
transportation. Despite the increasing need for situational awareness of
space objects and their re-entry, the classical risk avoidance paradigm
revolves around restricting the airspace. 
Such airspace closures impact the flow of air traffic, as the operating schedule will get distorted due to delays or re-routings circumnavigating the exposed volume of airspace.
While the probability of a strike is low, it can result in catastrophic consequences. 
The growing number of space operations requires therefore tools to help improve the monitoring and trajectory prediction of uncontrolled re-entries, techniques to help reduce the uncertainty in terms of the dimension and duration of required airspace measures, including computationally efficient and fast methods to determine the risk exposure.

@@

::: {#fig-cells}
```{r cells-map}
#| out-width: "120%"
#| echo: false
#| fig-scap: "H3 cells at resolution 3"
knitr::include_graphics("media/figures/cells_3.png")
```

@@

::: {#fig-density-summer2024-res-3}
```{r density-winter-res-3}
#| out-width: "110%"
#| width: "9cm"
#| echo: false
knitr::include_graphics("media/figures/flight_density_2024-07-05_3.png")
```

@@

The impact probability density $\sigma(\varphi, i)$ for an inclination $i$ of
$\ang{51.7}$ is plotted in @fig-weighting-inclination-i and takes a characteristic U shape with zero values for latitudes beyond $i$.
This density has been derived by simulating a full circular orbit at
\qty{550}{km} height with 10000  positions and calculating the relevant longitudes and latitudes at half degree resolution.

::: {#fig-weighting-inclination-i}
```{r weighting-inclination-i}
#| out-width: "100%"
#| echo: false
knitr::include_graphics("media/figures/weighting_inclination_i.png")
```

@@

@@

In few days (at the time of writing of this paper) at 23:06 May 10, 2025
($\pm 1.8$ days), the KOSMOS 482 Descent Craft should re-enter the
Earth's atmosphere [@kosmos482reentry].
This is the lander module from a 1972 failed USSR Venera mission [@SpaceReviewKosmos],
it is reported to weigh about $\qty{480}{kg}$ and to have a diameter of approximately $\qty{1}{m}$.
Being built to survive Venus's atmosphere which is much denser and hotter than Earth's one (surface pressure 93 times that of Earth and temperatures reaching \ang{467}C (\ang{872}F)),
it could re-enter Earth intact and not break into smaller fragments some of which could burn up.

::: {#fig-kosmos-cell}
```{r kosmos-cell}
#| out-width: "45%"
#| echo: false
knitr::include_graphics(here::here("media", "figures", "cell_london.png"))
```

Cell with highest collision expectation due to KOSMO debris re-entry based on 2024-07-05 traffic.
:::

Knowing that the debris inclination is $\ang{51.7}$, see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a proxy for the traffic density on the day of re-entry.
Using the approach in this study and the weightings for this specific debris,
the cell with the highest impact risk is cell, `83194afffffffff` (located South
of London, cf. @fig-kosmos-cell), with a value of $1.89 \cdot  10^{-10}$.

@@

We have calculated collision risk probabilities for the two
busiest weeks in 2023 and 2024. The relevant dataset represents a data cube
as logically depicted in @fig-data-densities.

@@

Code and sample data is available in the companion GitHub repository of
the paper at <https://github.com/euctrl-pru/flight_density_space_debris>.
We have also developed an R package `aviodebris` to heavy-lift the
various data processing steps to produce the results in the paper;
it is available at <https://github.com/euctrl-pru/aviodebris>.

::: {#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="htb"}
```{r ellipse-parameters}
#| fig-cap: "Keplerian elements: semi-major axis, $a$, the distance between the foci, $2c$, the eccentricity, $e = (2c)/(2a)$. (the semi-parameter (a.k.a. _semilatus rectum_), $p$, describes the size of the conic section by defining the width at the primary focus, and is defined as $p = b^2 / a = a (1 - e^2)$)"

@@

The findings are applied to a use-case: the imminent re-entry of the
lander module of an earlier Venus mission. The maximum collision
expectation for the busiest day in 2024 is determined.

@@

These are used to derive numerically compute the weighting function,
@fig-weighting-inclination-i:
for debris with inclination $i$ we simulate an orbit of $N$, i.e. $10001$,
positions and calculate the corresponding Cartesian position and the
equivalent longitude and latitude.

>>>>

format:
  ieee-pdf:
    keep-tex: true
    # fig-pos: 'H'
    journaltype: conference
    papersize: a4
    include-in-header:
      - file: media/preamble.tex
    include-before-body:
      - text: |
          \let\qty\SI
          % NOTE: make siunitx's \qty work without clashing with the one in physics
  ieee-html: 
    html-math-method: mathjax
    include-in-header:
      file: media/siunitx.html
author:
  - id: es
    name: Author 1
    affiliations:
      - ref: fake
    # name: Enrico Spinielli
    # affiliations:
    #   - ref: aiu
    # orcid: 0000-0001-8584-9131
    # email: enrico.spinielli@eurocontrol.int
    # attributes:
    #   corresponding: true
  - name: Author 2
    affiliations:
      - ref: fake
  # - name: Rainer Koelle
  #   affiliations: 
  #     - ref: aiu
  #   orcid: 0000-0003-2810-1404
  #   email: rainer.koelle@eurocontrol.int
  - name: Author 3
    affiliations:
      - ref: fake
  # - name: Quinten Goens
  #   affiliations: 
  #     - ref: aiu
  #   orcid: 0000-0003-2187-1924
  #   email: quinten.goens@eurocontrol.int
affiliations:
  - id: aiu
    name: EUROCONTROL
    department: Aviation Intelligence Unit
    city: Brussels
    country: Belgium
  - id: fake
    name: ACME
    department: Best Tricks
    city: Los Angeles
    country: USA
bibliography: media/references.bib
acronyms:
  insert_loa: false
  fromfile:
    - ./media/acronyms.yml
editor: 
  markdown: 
    wrap: 72
---
```{r}
```

@@

The recent decades observed an increase in the number of space activities.
This resulted in higher numbers of re-entries at the end of the respective 
mission lifespan or orbital decay. The risk of collision between uncontrolled
re-entering space debris and aircraft is increasing due to the continual 
growth of flights and of orbiting bodies and their relevant re-entries. 
<!-- why important --> National and international
authorities need to balance safety and economic impact of adopted
measures. This is particularly challenging should a re-entry event be announced on short notice.
Thus, there is a need for effective tools to reduce the uncertainty and increase the safety awareness of uncontrolled re-entries. 
<!-- approach / experiments --> To determine the risk exposure, this paper follows a data-driven approach.
We employ hourly and daily spatial flight densities calculated from trajectories
as recorded by EUROCONTROL's Network Manager and binned with the Uber H3
geospatial indexing system.
Impact risk probabilities are then calculated based on previous theoretical
work. 
<!-- results / conclusions --> This work enhances the previou model by using
actual flown flight trajectories as a basis for time-varying flight densities
rather than relying on proxies such as the population density or ADS-B/Mode S
message counts. These densities combined with a buffered area along the eventual path of re-entry, can provide mid-air impact probabilities to authorities
as an aid for deciding on safety measures such as airspace closures and/or
re-routings.
The findings and the approach are demonstrated on the basis of an expected re-entry from an earlier space mission.\
The approach in this paper enables a computing time efficient mechanism to support the re-entry related decision-making.

@@

With forecasts predicting a steadily growing number of flights, see
@fig-forecast, and the modern race to commercialise space, there is a
need to re-evaluate the risks of collision and casualties due to
uncontrolled debris re-entries. The increasing risk landscape poses
challenges in terms of oversight and risk management. The operational and 
economic demand by air transport requires a trade-off with large-scale
airspace and flow management measures to address the inherent safety risk.
There is a growing demand by overseeing authorities for methodologies and
tools to assess which safety measures are the most adequate to minimise
potential casualties and economic impact.

The increase in space operations does not exist in a vacuum. The
continual growth of air traffic - despite temporal bumps - is evidenced
by the surging number of flights in Europe [@forecastUpdate2025] and
across the world. On-going development/deployment activities of
mega-constellations of satellites in \acr{LEO}, such as the
already active Starlink and OneWeb or planned ones (e.g. Kuiper,
Starshield, Xingwang), will add thousand of orbiting objects,
c.f. [@tbl-megaconstellations;@megaconstellations] with a short
operational life. Some of these will inevitably re-enter the Earth's
atmosphere in an uncontrolled manner.
These numbers and occurrence rates will pose challenges to the steadily increasing air traffic.

@@

Aviation safety risk management is a well-established domain in air
transportation. Despite the increasing need for situational awareness of
space objects and their re-entry, the classical risk avoidance paradigm
revolves around restricting the airspace. 
Such airspace closures impact the flow of air traffic, as the operating schedule
will get distorted due to delays or re-routings circumnavigating the exposed volume of airspace.
While the probability of a strike is low, it can result in catastrophic 
consequences. 
The growing number of space operations requires therefore tools to help improve
the monitoring and impact of uncontrolled re-entries, techniques to help reduce
the uncertainty in terms of the dimension and duration of required airspace
measures, including computationally efficient and fast methods to determine the risk exposure.

@@

::: {#fig-cells}
```{r cells-map}
#| out-width: "110%"
#| echo: false
#| fig-scap: "H3 cells at resolution 3"
knitr::include_graphics("media/figures/cells_3.png")
```

@@

::: {#fig-density-summer2024-res-3}
```{r density-winter-res-3}
#| out-width: "105%"
#| width: "9cm"
#| echo: false
knitr::include_graphics("media/figures/flight_density_2024-07-05_3.png")
```

@@

::: {#fig-weighting-inclination-i}
```{r weighting-inclination-i}
#| out-width: "90%"
#| echo: false
knitr::include_graphics("media/figures/weighting_inclination_i.png")
```

@@

The impact probability density $\sigma(\varphi, i)$ for an inclination $i$ of
$\ang{51.7}$ is plotted in @fig-weighting-inclination-i and takes a characteristic U shape with zero values for latitudes beyond $i$.
This density has been derived by simulating a full circular orbit as detailed
in @#sec-orbitals.

@@

In few days (at the time of writing of this paper) at 10:05 May 10, 2025
($\pm 1.5$ days), the KOSMOS 482 Descent Craft should re-enter the
Earth's atmosphere [@kosmos482reentry].
This is the lander module from a 1972 failed USSR Venera mission [@SpaceReviewKosmos],
it is reported to weigh about $\qty{480}{kg}$ and to have a diameter of approximately $\qty{1}{m}$.
Being built to survive Venus's atmosphere which is much denser and hotter than Earth's one (surface pressure 93 times that of Earth and temperatures reaching \ang{467}C (\ang{872}F)),
it could re-enter Earth intact and not break into smaller fragments some of which could burn up.

Knowing that the debris inclination is $\ang{51.7}$, see @fig-weighting-inclination-i, we can derive the maximum collision expectation using data for the busiest day in the busiest 2024 week as a proxy for the traffic density on the day of re-entry.
Using the approach in this study and the weightings for this specific debris,
the cell with the highest impact risk is cell, `83194afffffffff` (located South of London), with a value of $1.89 \cdot  10^{-10}$.

@@

We have calculated collision risk probabilities for the two
busiest weeks in 2023 and 2024. The relevant dataset represents a data cube
as logically depicted in @fig-data-densities.
The processing of data for 1 day takes less than 30 min from
data extraction to impact probability calculation on a
high-end laptop (MacBook Pro M1 Max).

@@

Code and sample data are available in the companion GitHub repository of
the paper at <https://github.com/euctrl-pru/flight_density_space_debris>.
We have also developed an R package `aviodebris` to heavy-lift the
various data processing steps which  is available at
<https://github.com/euctrl-pru/aviodebris>.

::: {#fig-orbital-elements layout="[40,-4,56]" fig-env="figure*" fig-pos="h"}
```{r ellipse-parameters}
#| fig-cap: "Keplerian elements: semi-major axis, $a$, the distance between the foci, $2c$, the eccentricity, $e = (2c)/(2a)$. (the semi-parameter (a.k.a. _semilatus rectum_), $p$, describes the size of the conic section by defining the width at the primary focus, and is defined as $p = b^2 / a = a (1 - e^2)$)"

@@

The findings are applied to a use-case: the imminent re-entry of the
lander module of an earlier failed Venus mission. The maximum collision
expectation for the busiest day in 2024 is determined.

@@

These are used to derive numerically compute the weighting function,
@fig-weighting-inclination-i:
for debris with inclination $i$ we simulate an orbit of $N$, i.e. $10000$,
positions and calculate the corresponding Cartesian position and the
equivalent longitude and latitude.
